\texttt{gwsnr}: A Python Package for Efficient SNR Calculation of GWs

### 2024-11-07 arxiv/paper.tex
mac

Gravitational waves (GWs), ripples in spacetime predicted by Einstein's
theory of General Relativity, have revolutionized astrophysics since
their first detection in 2015 (Abbott, B. P. et al. (2016a)). These
waves are emitted by cataclysmic events like the merging of binary black
holes (BBHs), binary neutron stars (BNSs) and BH-NS pairs, providing a
unique window into the cosmos. A critical aspect of GW analysis is the
Signal-to-Noise Ratio (SNR). SNR quantifies the strength of a GW signal
relative to the background noise in a detector, like LIGO (The LIGO
Scientific Collaboration et al. (2015)) or Virgo (Acernese et al.
(2014)). This ratio is pivotal in confirming the detection of GWs and
extracting astrophysical information from them (Abbott, B. P. et al.
(2016b)). However, specific scenarios in GW research, particularly in
simulations of detectable GW events (Abbott et al. (2016)) and in
hierarchical Bayesian analysis (Thrane and Talbot (2019)) where
selection effects are considered, demand extensive and efficient
computation of SNR. This requirement presents a significant challenge,
as conventional computational approaches, such as noise-weighted inner
product, are typically time-consuming and impractical for such
specialized and large-scale analyses (Taylor and Gerosa (2018), Davide,
Pratten, and Vecchio (2020)).

@@

The \emph{\texttt{gwsnr}} package now incorporates an artificial neural
network (ANN) model, developed using \emph{\texttt{TensorFlow}} (Abadi
et al. (2015)) and \emph{\texttt{sklearn}} (Pedregosa et al. (2011)), to
rapidly estimate the Pdet in binary black hole (BBH) systems using the
IMRPhenomXPHM waveform approximant. This complex IMR waveform model
accounts for spin-precessing systems with subdominant harmonics. The ANN
model is especially useful when precise signal-to-noise ratio (SNR)
calculations are not critical, providing a quick and effective means of
estimating Pdet. This value indicates detectability under Gaussian noise
by determining if the SNR exceeds a certain threshold. Trained on a
large dataset from the \emph{\texttt{ler}} package, the ANN model uses
`partial scaled SNR' values as a primary input, reducing input
dimensionality from 15 to 5 and enhancing accuracy. This approach offers
a practical solution for assessing detectability under specified
conditions.

@@

\bibitem[\citeproctext]{ref-Gerosa:2020}
Davide, Gerosa, Geraint Pratten, and Alberto Vecchio. 2020.
{``Gravitational-Wave Selection Effects Using Neural-Network
Classifiers.''} \emph{Physical Review D} 102 (10).
\url{https://doi.org/10.1103/physrevd.102.103020}.

>>>>

Gravitational waves (GWs), ripples in spacetime predicted by Einstein's
theory of General Relativity, have revolutionized astrophysics since
their first detection in 2015 (Abbott, B. P. et al. (2016a)). These
waves are emitted by cataclysmic events like the merging of binary black
holes (BBHs), binary neutron stars (BNSs) and BH-NS pairs, providing a
unique window into the cosmos. A critical aspect of GW analysis is the
Signal-to-Noise Ratio (SNR). SNR quantifies the strength of a GW signal
relative to the background noise in a detector, like LIGO (The LIGO
Scientific Collaboration et al. (2015)) or Virgo (Acernese et al.
(2014)). This ratio is pivotal in confirming the detection of GWs and
extracting astrophysical information from them (Abbott, B. P. et al.
(2016b)). However, specific scenarios in GW research, particularly in
simulations of detectable GW events (Abbott et al. (2016)) and in
hierarchical Bayesian analysis (Thrane and Talbot (2019)) where
selection effects are considered, demand extensive and efficient
computation of SNR. This requirement presents a significant challenge,
as conventional computational approaches, such as noise-weighted inner
product, are typically time-consuming and impractical for such
specialized and large-scale analyses (Taylor and Gerosa (2018), Gerosa
et al. (2020)).

@@

The \emph{\texttt{gwsnr}} package now incorporates an artificial neural
network (ANN) model, developed using \emph{\texttt{TensorFlow}} (Abadi
et al. (2015)) and \emph{\texttt{sklearn}} (Pedregosa et al. (2011)), to
rapidly estimate the Pdet in binary black hole (BBH) systems using the
IMRPhenomXPHM waveform approximant. This complex IMR waveform model
accounts for spin-precessing systems with subdominant harmonics. The ANN
model is especially useful when precise signal-to-noise ratio (SNR)
calculations are not critical, providing a quick and effective means of
estimating Pdet. This value indicates detectability under Gaussian noise
by determining if the SNR exceeds a certain threshold. Trained on a
large dataset from the \emph{\texttt{ler}} package, the ANN model uses
`partial scaled SNR' values as a primary input, reducing input
dimensionality from 15 to 5 and enhancing accuracy. This approach offers
a practical solution for assessing detectability under specified
conditions. Other similar efforts with ANN models are detailed in
Chapman-Bird et al. (2023), Gerosa et al. (2020) etc.

@@

\bibitem[\citeproctext]{ref-ChapmanBird2023}
Chapman-Bird, Christian E A et al. 2023. {``Rapid Determination of LISA
Sensitivity to Extreme Mass Ratio Inspirals with Machine Learning.''}
\emph{Monthly Notices of the Royal Astronomical Society} 522 (4):
6043--54. \url{https://doi.org/10.1093/mnras/stad1397}.

\bibitem[\citeproctext]{ref-Gerosa:2020}
Gerosa, Davide et al. 2020. {``Gravitational-Wave Selection Effects
Using Neural-Network Classifiers.''} \emph{Physical Review D} 102 (10).
\url{https://doi.org/10.1103/physrevd.102.103020}.
